### 2020-03-03 doc/icfp20/counterpoint.tex
improve comparison section

The type-theoretical formalization of counterpoint has previously
been attempted by \citet{cong-cp}.
They represent correct counterpoint as an Agda datatype akin to
a list, where the base cases enforce a valid ending and the inductive
case guarantees correct uses of intervals and motion (by means of
implicit arguments representing proofs).

Our representation of counterpoint improves on the existing
implementation in two ways.
First, while \citet{cong-cp} rely on the Agda type checker to report
errors in the counterpoint, we use our own type checker (implemented
as the \texttt{checkXXX} functions) and type errors (defined as the
\texttt{XXXError} datatypes).
This allows us to produce error messages that clearly state what
kind of mistake the programmer has made.
Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
an octave (8th) of D, breaking the rule that perfect intervals cannot
be approached by parallel motion.
In our implementation, this causes an error in the \texttt{motionOk}
field, with the following message:

@@

\noindent In \citet{cong-cp}'s implementation, on the other hand,
the invalid motion results in an unsolved metavariable \texttt{\_13},
meaning that Agda failed to construct a proof required by the
counterpoint constructor:

@@

\noindent For a non-expert user, it may be difficult to connect
this message to the corresponding musical error \footnote{
  The attempt to produce customized musical errors has also
  appeared in \citet{szamozvancev-welltyped}.
  They formalize various musical rules, including those for
  intervals and motion, using GHC's custom type error feature.
  However, their error messages are not as precise as ours:
  e.g., for the parallel octave example, they simply report
  ``\texttt{Direct motion into a perfect octave is forbidden}'',
  without the precise information about the two intervals and
  the motion between them.}.

>>>>

The type-theoretical formalization of counterpoint has first
been attempted by \citet{szamozvancev-welltyped}.
They encode the rules for intervals and motion by declaring
type classes \texttt{ValidHarmInterval} and \texttt{ValidMotion},
and specifying which instances are invalid using GHC's support
for custom type errors.

The availability of the custom error feature is an advantage of using
Haskell, but the language's type system does not allow producing
detailed error messages.
Suppose we have replaced the second interval of Figure~\ref{fig:fs} by
an octave (8th) of D, breaking the rule that perfect intervals cannot
be approached by parallel motion.
In our implementation, this causes an error in the \texttt{motionOk}
field, with the following message:

@@

\noindent Observe that the message clearly tells us which intervals
form what kind of invalid motion.

In Szamozvancev and Gale's implementation, the same example
results in the following error:

\begin{alltt}
Direct motion into a perfect octave is forbidden.
\end{alltt}

\noindent The lack of the precise information about the intervals
and motion is due to the absence of native dependent types in
Haskell.

More recently, \citet{cong-cp} give an implementation of first
species counterpoint using dependent types in Agda.
Their idea is to represent correct counterpoint as a list-like
datatype, where the base cases enforce a valid ending and the
inductive case guarantees correct uses of intervals and motion
(by means of implicit arguments representing proofs).

Our representation of counterpoint improves on their
implementation in two ways.
First, while \citet{cong-cp} rely on the Agda type checker to report
errors in the counterpoint, we use our own type checker (implemented
as the \texttt{checkXXX} functions) and type errors (defined as the
\texttt{XXXError} datatypes).
This allows us to produce helpful error messages while keeping
types simple.
Indeed, for the parallel octave example discussed above,
\citet{cong-cp} would just report the existence of an unsolved
metavariable \texttt{\_13}, meaning that Agda failed to construct
a proof required by the counterpoint constructor:

@@

\noindent For a non-expert user, it may be difficult to connect
this message to the corresponding musical error.
